---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# incR: R package for the analysis of animal incubation

This is a beta version of an R package developed to analyse animal incubation. 
The code and the documentation are being actively updated. 

incR takes a time-series of incubation temperature, scores incubation and provides
a suite of functions to calculate biologically relevant metrics.

Please, if you use the package and find bugs or think of any improvement, get in 
contact to implement your recommendations. 


## Installation

You can install incR from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("PabloCapilla/incR")
```

Additionally, the package will be soon submitted to CRAN.

[![Travis-CI Build Status](https://travis-ci.org/PabloCapilla/incR.svg?branch=master)](https://travis-ci.org/PabloCapilla/incR)