
<!-- README.md is generated from README.Rmd. Please edit that file -->

[![Travis-CI Build Status](https://travis-ci.org/PabloCapilla/incR.svg?branch=master)](https://travis-ci.org/PabloCapilla/incR)

# incR: R package for the analysis of animal incubation

This is the development site of `incR`, an R package to analyse avian incubation temperatures and
incubation behaviour. 
The code and the documentation are updated often - check for latest version here and on CRAN.

`incR` takes a time-series of incubation temperature, scores incubation and provides
a suite of functions to calculate biologically relevant metrics of incubation.

Please, if you use the package and find bugs or think of any improvement, get in 
contact to implement your recommendations. 


## Installation

You can install `incR` from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github(repo = "PabloCapilla/incR")
```
Or from CRAN:
```{r eval = FALSE}
install.packages("incR")
```

Read the package's vignettes to find out how it works!
```{r eval = FALSE}
browseVignettes("incR")
``` 

## Citation

Please, if you use `incR` in your publications cite it as:
```{r eval = TRUE}
citation("incR")
``` 